Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis

### 2024-12-06 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update prose issue

Our Bayesian DID model offers a robust approach to examining the impact
of abortion bans on infant mortality but has limitations. Misspecified
priors, data imbalances, and violations of the parallel trends
assumption could introduce biases. Despite these challenges, the model
effectively integrates prior knowledge and addresses uncertainty,
providing a valuable understanding of policy impacts.

@@

The analysis reveals notable patterns in infant mortality rates
influenced by policy and demographic factors. For states where abortion
remained legal, the post-injunction period (after_injunction1) is
associated with reductions in infant mortality rates, with a decrease of
0.525 deaths per 1,000 live births in the first model and 0.947 in the
second model, compared to the pre-injunction period. In contrast, states
where abortion became illegal (abortion_illegal1) experienced
significantly higher infant mortality rates, with increases of 3.365
deaths in the first model and 1.401 in the second model, relative to
states where abortion remained legal.

@@

The inclusion of control variables in the second model significantly improves its explanatory power compared to the first model, as evidenced by the higher adjusted $R^2$ (0.778) in the second model versus 0.213 in the first. These control variables—such as maternal age, race, and year of death—capture critical demographic and temporal variations influencing infant mortality rates, leading to more precise and reliable estimates of the impact of abortion bans.

@@

The results reveal stark racial disparities in infant mortality rates,
with Black mothers experiencing significantly higher rates compared to
other racial groups. These disparities are exacerbated by restrictive
abortion policies, which disproportionately affect already vulnerable
populations. The intersection of systemic healthcare inequities,
socioeconomic challenges, and restrictive legislation creates compounded
risks for marginalized communities, leading to worse health outcomes.
Addressing these inequities requires targeted policies that ensure
equitable access to maternal and infant healthcare, particularly in
communities most impacted by these disparities. Such policies should
prioritize investments in healthcare infrastructure, culturally
competent care, and programs that support at-risk populations.

@@

The RMSE values from cross-validation indicate that the first model and the second model perform similarly in predicting infant mortality rates,
with RMSE values of 5.07 and 5.06, respectively. This similarity
suggests that adding demographic factors such as maternal age and race
in the second model has only a marginal impact on predictive accuracy
compared to the first model, which primarily relies on policy-related
variables like abortion legality and timing of the injunction. An average error of approximately 5 deaths per 1,000 live births indicates
the magnitude of prediction errors, which may or may not be acceptable
depending on the variability in the dataset. While the second model
slightly reduces the error, its added complexity may not justify the
trade-off if simplicity is a priority. However, if a nuanced
understanding of demographic influences is essential, the second model
provides a modest improvement in accuracy while offering more detailed
insights into maternal and infant health disparities. The
cross-validation process reinforces the reliability of these models by
confirming their performance across multiple data splits, reducing the
risk of overfitting.

@@

Introduction: This section seeks to understand the experiences and
outcomes of women who, due to restrictive abortion laws, were unable to
access abortion services and subsequently faced the loss of their child.
We recognize that discussing past pregnancies can be sensitive,
especially those involving circumstances such as abortion or the loss of
a child. Your understanding are invaluable in identifying areas where
healthcare and support services can be improved. If you prefer not to
answer, you are welcome to skip this section.

>>>>

Our Bayesian DID model offers a robust approach to examining the impact
of abortion bans on infant mortality but has limitations. Misspecified
priors, data imbalances, and violations of the parallel trends
assumption could introduce biases. Despite these challenges, the model
effectively integrates prior knowledge and addresses uncertainty,
providing an important understanding of policy impacts.

@@

The analysis explains notable patterns in infant mortality rates
influenced by policy and demographic factors. For states where abortion
remained legal, the post-injunction period (after_injunction1) is
associated with reductions in infant mortality rates, with a decrease of
0.525 deaths per 1,000 live births in the first model and 0.947 in the
second model, compared to the pre-injunction period. In contrast, states
where abortion became illegal (abortion_illegal1) experienced
significantly higher infant mortality rates, with increases of 3.365
deaths in the first model and 1.401 in the second model, relative to
states where abortion remained legal.

@@

The inclusion of control variables in the second model significantly improves its explanatory power compared to the first model, as evidenced by the higher adjusted $R^2$ (0.778) in the second model versus 0.213 in the first. These control variables—such as maternal age, race, and year of death—capture detailed demographic and temporal variations influencing infant mortality rates, leading to more precise and reliable estimates of the impact of abortion bans.

@@

The results explain stark racial disparities in infant mortality rates,
with Black mothers experiencing significantly higher rates compared to
other racial groups. These disparities are exacerbated by restrictive
abortion policies, which disproportionately affect already vulnerable
populations. The intersection of systemic healthcare inequities,
socioeconomic challenges, and restrictive legislation creates compounded
risks for marginalized communities, leading to worse health outcomes.
Addressing these inequities requires targeted policies that ensure
equitable access to maternal and infant healthcare, particularly in
communities most impacted by these disparities. Such policies should
prioritize investments in healthcare infrastructure, culturally
competent care, and programs that support at-risk populations.

@@

The RMSE values from cross-validation indicate that the first model and the second model perform similarly in predicting infant mortality rates,
with RMSE values of 5.07 and 5.06, respectively. This similarity
suggests that adding demographic factors such as maternal age and race
in the second model has only a marginal impact on predictive accuracy
compared to the first model, which primarily relies on policy-related
variables like abortion legality and timing of the injunction. An average error of approximately 5 deaths per 1,000 live births indicates
the magnitude of prediction errors, which may or may not be acceptable
depending on the variability in the dataset. While the second model
slightly reduces the error, its added complexity may not justify the
trade-off if simplicity is a priority. However, if a detailed understanding of demographic influences is essential, the second model
provides a modest improvement in accuracy while offering a more detailed understanding of maternal and infant health disparities. The
cross-validation process reinforces the reliability of these models by
confirming their performance across multiple data splits, reducing the
risk of overfitting.

@@

Introduction: This section seeks to understand the experiences and
outcomes of women who, due to restrictive abortion laws, were unable to
access abortion services and subsequently faced the loss of their child.
We recognize that discussing past pregnancies can be sensitive,
especially those involving circumstances such as abortion or the loss of
a child. Your understanding is invaluable in identifying areas where
healthcare and support services can be improved. If you prefer not to
answer, you are welcome to skip this section.
